---
title: "FINAL"
author: "Maxi Ahl"
date: "2026-02-05"
output: html_document
---

```{r,silent=TRUE}
library(terra)      # Reading and manipulating raster data
library(sf)         # Reading and manipulating vector data
library(dplyr)      # Manipulating data.frames
library(ggplot2)    # Visualization
library(spatstat)   # Point Pattern Analysis
library(ggplot2)    # data visualizatio
library(tidyr)      # data wrangling package: change from wide to long format
library(dplyr)      # data manipulation: mutate
library(gridExtra)  # plot multiple ggplot graphs
library(caret)      # cross-validation
library(mapview)    # Map visualization
library(gstat)      # geostatistics
library(stars)      # spatial-temporal data
library(abind)
library(spdep)      # Diagnosing spatial dependence
library(spatialreg) # Spatial lag and spatial error model
library(readxl)
library(raster)
library(ggcorrplot)
library(reshape2)
library(pracma)
library(data.table)
library(stringr)
library(readr)
library(corrplot)
library(broom)
library(spatialreg)
```


LOAD IN FINAL GPKG
```{r}
FULL <- st_read("/Users/maxi_161/Desktop/UNI/Master/THESIS/DATEN + GIS/ANALYSIS.nosync/gemeinden_elbe_final_full.gpkg")

glimpse(FULL)
```

CHECKS
Protection
```{r}
summary(FULL$share_protected)
```

Exposure
```{r}
summary(FULL$exposure_share_flood_total)
```

Kombination
```{r}
cor(FULL$share_protected,
    FULL$exposure_share_flood_total,
    use="complete.obs")
```

Residual Risk
```{r}
FULL <- FULL %>%
  mutate(
    residual_risk =
      exposure_share_flood_total * (1 - share_protected)
  )
```

```{r}
summary(FULL$residual_risk)
```

```{r}
head(FULL)
```

```{r}
vars <- FULL |>
  sf::st_drop_geometry() |>
  dplyr::select(
    exposure_share_alg2_sgb2,
    exposure_share_sgb2_with_housing_costs,
    exposure_share_longterm_unemp,
    exposure_unemp_u25_per_1000,
    exposure_purchasing_power,
    exposure_tax_revenue_total,
    exposure_share_age_65plus,
    exposure_share_age_0_3,
    exposure_natural_pop_change,
    exposure_migration_balance,
    exposure_old_age_dependency,
    exposure_share_single_households,
    exposure_share_hh_income_low,
    exposure_students_total_per_1000,
    exposure_doctors_total_per_1000,
    exposure_gp_primary_per_1000,
    exposure_share_bb_100mbit,
    exposure_share_4g,
    exposure_dist_public_transport_m,
    exposure_dist_pharmacy_m,
    exposure_pop_density_per_km2
  ) |>
  dplyr::mutate(
    dplyr::across(dplyr::everything(), ~ readr::parse_number(as.character(.x)))
  )

cor_mat <- cor(vars, use = "pairwise.complete.obs")
corrplot::corrplot(cor_mat, method = "color", type = "upper", tl.cex = 0.7)
```



RESIDUAL RISK CONCEPT
Zone 1 = offizielles Überschwemmungsgebiet

→ regelmäßig betroffen
→ gesetzlich festgesetzt
→ ohne Schutzmaßnahme

Zone 3 = hochwassergeschütztes Gebiet

→ liegt hinter Deichen
→ wäre ohne Deich im Überflutungsgebiet
→ wird aber in Extremereignissen modelliert

Unprotected flood exposure   = zone1
Protected flood exposure     = zone3
Total flood exposure         = zone1 + zone2 + zone3


# 1) Make sure zone shares exist and NAs are 0

```{r}
FULL <- FULL %>%
  mutate(
    exposure_share_flood_total = if_else(is.na(exposure_share_flood_total), 0, exposure_share_flood_total),
    exposure_share_flood_zone1 = if_else(is.na(exposure_share_flood_zone1), 0, exposure_share_flood_zone1),
    exposure_share_flood_zone2 = if_else(is.na(exposure_share_flood_zone2), 0, exposure_share_flood_zone2),
    exposure_share_flood_zone3 = if_else(is.na(exposure_share_flood_zone3), 0, exposure_share_flood_zone3)
  )
```

# 2) Define core outcomes
```{r}
FULL <- FULL %>%
  mutate(
    share_total_hazard = exposure_share_flood_total,
    share_unprotected  = exposure_share_flood_zone1,
    share_informational = exposure_share_flood_zone2,
    share_protected_hazard = exposure_share_flood_zone3,

    # Residual risk (supervisor-consistent): what is unprotected in HQ100
    residual_risk = share_unprotected,

    # How much of the hazard footprint is classified as protected (avoid /0)
    protection_share_of_hazard = if_else(
      share_total_hazard > 0,
      share_protected_hazard / share_total_hazard,
      NA_real_
    )
  )

```

# 3) Sanity checks
```{r}
summary(FULL$share_total_hazard)
```

```{r}
summary(FULL$share_unprotected)
```

```{r}
summary(FULL$share_protected_hazard)
```

```{r}
summary(FULL$protection_share_of_hazard)
```

# Optional quick check: do zone parts roughly add up to total?
```{r}
check_diff <- FULL %>%
  st_drop_geometry() %>%
  transmute(
    diff = share_total_hazard - (share_unprotected + share_informational + share_protected_hazard)
  )

summary(check_diff$diff)
```


WIDE PCA (Principal Component Analysis)
# 1. Extract all socio-economic vars
```{r}
names(FULL)
```

```{r}
wide_pca_vars <- FULL %>%
  sf::st_drop_geometry() %>%
  dplyr::select(

    # Poverty / Welfare
    exposure_share_alg2_sgb2,
    exposure_share_bg_single_parent,
    exposure_share_bg_5plus,
    exposure_share_bg_with_children,
    exposure_share_sgb2_with_housing_costs,

    # Unemployment
    exposure_share_longterm_unemp,
    exposure_unemp_u25_per_1000,
    exposure_unemp_55plus_per_1000,

    # Income / Fiscal
    exposure_income_tax_per_capita,
    exposure_purchasing_power,
    exposure_trade_tax_per_capita,
    exposure_tax_revenue_total,

    # Demography
    exposure_share_age_0_3,
    exposure_share_age_3_6,
    exposure_share_age_6_18,
    exposure_share_age_18_25,
    exposure_share_age_25_30,
    exposure_share_age_30_50,
    exposure_share_age_50_65,
    exposure_share_age_65_75,
    exposure_share_age_65plus,
    exposure_share_age_75plus,

    # Population dynamics
    exposure_natural_pop_change,
    exposure_migration_balance,

    # Dependency
    exposure_old_age_dependency,
    exposure_youth_dependency,
    exposure_young_old_ratio,

    # Household
    exposure_share_single_households,
    exposure_share_households_with_children,

    # Income / education
    exposure_share_hh_income_high,
    exposure_share_hh_income_medium,
    exposure_share_hh_income_low,
    exposure_students_total_per_1000,
    exposure_students_18_25_per_1000,
    exposure_students_fh_per_1000,

    # Health
    exposure_gp_general_per_1000,
    exposure_gp_primary_per_1000,
    exposure_internists_per_1000,
    exposure_pediatricians_per_1000_children,
    exposure_doctors_total_per_1000,

    # Digital
    exposure_share_bb_1000mbit,
    exposure_share_bb_100mbit,
    exposure_share_bb_50mbit,
    exposure_share_4g,

    # Accessibility
    exposure_dist_supermarket_m,
    exposure_dist_pharmacy_m,
    exposure_dist_gp_m,
    exposure_dist_public_transport_m,
    exposure_dist_primary_school_m,

    # Density
    exposure_pop_density_per_km2,
    exposure_employment_density_per_km2
  ) %>%
  mutate(across(everything(), as.numeric)) %>%
  rename_with(~ stringr::str_remove(.x, "^exposure_"))
```

```{r}
names(wide_pca_vars)
```

```{r}
wide_pca_clean <- wide_pca_vars %>% drop_na()
wide_pca_scaled <- scale(wide_pca_clean)

wide_pca <- prcomp(wide_pca_scaled)

summary(wide_pca)
round(wide_pca$rotation[,1:5], 3)
```


# ---- PCA Loadings Table ----
```{r}
loadings <- as.data.frame(wide_pca$rotation)
loadings$variable <- rownames(loadings)

# Long format
loadings_long <- loadings %>%
  pivot_longer(
    cols = starts_with("PC"),
    names_to = "PC",
    values_to = "loading"
  ) %>%
  mutate(abs_loading = abs(loading))

# Top 8 variables per PC
top_loadings <- loadings_long %>%
  group_by(PC) %>%
  arrange(desc(abs_loading)) %>%
  slice_head(n = 8) %>%
  ungroup()

# View
top_loadings %>%
  arrange(PC, desc(abs_loading))
```

# ---- Scree Plot ----

```{r}
pca_var <- wide_pca$sdev^2
pca_var_prop <- pca_var / sum(pca_var)

scree_df <- data.frame(
  PC = seq_along(pca_var),
  Eigenvalue = pca_var,
  Variance = pca_var_prop,
  Cumulative = cumsum(pca_var_prop)
)

ggplot(scree_df, aes(x = PC, y = Eigenvalue)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(
    title = "Scree Plot (Kaiser Criterion)",
    x = "Principal Component",
    y = "Eigenvalue"
  ) +
  theme_minimal()
```
Kaiser-Kriterium (Standard)
```{r}
scree_df %>%
  filter(Eigenvalue > 1)
```
Cumulative Variance Check
```{r}
scree_df %>%
  filter(Cumulative <= 0.7)
```


Hauptindex: PC1–PC8
Sensitivity: PC1–PC12

Hauptversion (PC1–PC8):

```{r}

```

```{r}
# --- PCA scores back into FULL (matches the drop_na rows) ---

# which rows were kept for PCA?
rows_kept <- complete.cases(wide_pca_vars)   # logical vector length nrow(FULL)

# create empty PC columns in FULL
k <- ncol(wide_pca$x)                        # number of PCs (51)
pc_names <- paste0("PC", 1:k)

FULL[, pc_names] <- NA_real_

# fill only the kept rows with the PCA scores
FULL[rows_kept, pc_names] <- wide_pca$x
```

```{r}
names(FULL)[grepl("^PC\\d+$", names(FULL))] |> head()
summary(FULL$PC1)
sum(!is.na(FULL$PC1))
```


INDEX BAUEN (Hauptindex PC1–PC8)
```{r}
k_main <- 8
weights_main <- scree_df$Variance[1:k_main]
names(weights_main) <- paste0("PC", 1:k_main)

pc_mat <- FULL |>
  sf::st_drop_geometry() |>
  dplyr::select(dplyr::all_of(names(weights_main))) |>
  as.matrix()

FULL$vuln_index_main <- as.numeric(pc_mat %*% weights_main)
summary(FULL$vuln_index_main)
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

